---
title: "Create TSD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create TSD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TimeSpatData)
```

The dimension `Spat` and `Time` based on the package `terra` and `lubridate`, so it's necessary to load this package:


```{r load}
library(terra)
library(lubridate)
```


# Create the TSD-data

## `TimeRastData` based on Raster data

There are two ways to create the `TimeRastData`, either read Raster-Cluster (multi-layers) into `array` or `terra::SpatRaster`:

### Load data

In this tutorial will convert three days meteorological data (potential ET) from [ReKIS](https://rekisviewer.hydro.tu-dresden.de/fdm/ReKISExpert.jsp) to TSD-data, the original data is saved as one of the raster format ASC-format.

```{r read2array}
## get the file path and names of raster data (in .asc)
fp_Extdata <- system.file("extdata", package = "TimeSpatData") 
fn_Asc <-  list.files(fp_Extdata, ".asc")

## read data into SpatRaster and array
rast_ET <- rast(file.path(fp_Extdata, fn_Asc[1:3]))  # the first six lines are the metadata
ary_ET <- as.array(rast_ET)

## check the data:
rast_ET 
str(ary_ET)
```

### From `array` to `TimeRastData`

The convention of TSD-Data define the **fest order of dimensions**, therefor the dimension of array must be consistent like aim TSD-Data stay.

The TSD-data anfordert always the geological information (`Spat`), but 
the `array` data class can't save any geological data, so we need still set the arguments `Spat_EPSG` and `Spat_extent`:

The dimension `Time` just need to convert the timestempal to the `Date` with `lubridate`

**NOTE: The TSD-Data define the y-Aches always from SMALL to BIG, but the most raster data save them from north to south, so the y-value ist from big to small. Therefor we need adjust this dimension.**

```{r array2trd}
## reform the array
ary_ET_reformed <- ary_ET[190:1,,] |> aperm(c(3,2,1))

## get the geological info
ext_Template <- ext(rast_ET)

## convert to TSD
tsd_ET_a <- new_TimeRastVariable(ary_ET_reformed, # data
                               "ET", "mm/d", # name and unit
                               as_date(c("1961-01-01", "1961-01-02", "1961-01-03")), # time
                               31468, ext_Template) # Spat

## check
str(tsd_ET_a)
```

This method is first recognized, because it can easy deal with the **huge** data.

### From `SpatRaster` to `TimeRastData`

The second method is only recognized for data that contain not so many time step in `Time` dimension.

```{r}
## add the crs infomation to the data, this depend on wether the raster-file contain this infomation, 
## when es contauin this info in original, then this step is not necessery
crs(rast_ET) <- "EPSG:31468"
## convert to TSD
tsd_ET_r <- new_TimeRastVariable(rast_ET, # data
                               "ET", "mm/d", # name and unit
                               as_date(c("1961-01-01", "1961-01-02", "1961-01-03"))) # time

## check
str(tsd_ET_r)

```



## `TimeVectData` based on Vector (and table) data

Compare to the raster data, the vector data save always only the geometric data, 
the other time-based data will be saved in separated data.
Therefor the both parties are always needed by the create of the `TimeVectData`.

### Load data

The polygons of Catchments "Mulde" and "Speree" in Saxony, German is download from [iDA (interdisziplinäre Daten und Auswertungen)](https://www.umwelt.sachsen.de/umwelt/infosysteme/ida/index.xhtml) and in GPKG-Format reformed.

```{r read2vect}
## load polygon
polygon_SAX <- vect(file.path(fp_Extdata, "shp_SAX_EZG.gpkg"))

## random array data
ary_Q <- array(runif(10, 0, 6), c(5, 2)) # random number for two `Spat` (polygons) and five `Time` (time steps)

```

### From `SpatVector` and `array` to `TimeVectData`

**NOTE: the Spat_Data MUST contain the attribute "Spat_ID"**

```{r}
## convert to TSD
tsd_Q_v <- new_TimeVectVariable(ary_Q, # data
                               "Q", "m3/s", # name and unit
                               as_date(c("1961-01-01", "1961-01-02", "1961-01-03", "1961-01-04", "1961-01-05")), # time
                               c("PID560051", "PID582820"), # Spat_ID
                               polygon_SAX) # Spat data, that must contain the attribute "Spat_ID"

## check
str(tsd_Q_v)

```






